---
title: "pgs-phenotype-analysis"
author: '`r Sys.info()["user"]`'
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: true
    highlight: tango
    code_folding: hide
params:
  input_phenotypes: "input/example.phenotypes.txt"
  input_PGS_data: "input/example.1000genomes.scores.txt"
  input_phenotypes_specification: "Phenotype_1,Phenotype_2,Phenotype_3"
  input_PGS_numbers: "PGS002695,PGS002695,PGS002695"
  input_variants_file: "input/variants.csv"
  input_remove_neg_values: TRUE
---


Version: 1.0

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r packages, echo = FALSE, include=F} 
library(tidyr) # data formatting
library(dplyr) # data manipulation
library(ggplot2) # various plots
library(DT) # datatable() for aesthetic formating of tables
library(patchwork) # combine plots into one image
library(latex2exp) # LaTeX expressions in plots
source("pgs-utils_jonas.R") # some functions for pgs-analysis in general
```

```{r definition of functions}
get_unique_number_func <- function(a_data, a_column){
  # a_column --> input as string! e.g "sample"
  count_value <- a_data %>% 
    select(all_of(a_column)) %>%
    unique() %>%
    nrow()
  return (count_value)
}



scatterplot_func <- function(a_data, a_score, a_phenotype){
  reg <- lm(get(a_phenotype)~get(a_score), data=a_data)
      reg_sum <- summary(reg)
      r2 <- round(reg_sum$r.squared, digits=3)
      
  plot <- ggplot(a_data, aes(y = get(a_phenotype), x = get(a_score))) +
    geom_point(color='lightblue', size = 1, alpha=0.4) + 
    ylab(a_phenotype) + xlab(a_score) +
    geom_smooth(method='lm', formula= y~x, color='blue', linewidth=0.5) +
    annotate(geom = 'text', label = paste("italic(R^2) ==", r2, " ") , parse=TRUE,  x = Inf, y = Inf, hjust = 1.1 , vjust = 1.1 ) +
    theme_bw()
  return(plot)
}


imputation_quality_histogram_func <- function(a_data, a_quality){
  quality_count <- a_data %>% filter(a_quality >= 0.9) %>% nrow()
  quality_count_form <- format(quality_count, big.mark= ",", scientific =FALSE)
  
  string_text <- paste(" $r^2 \\geq  0.9$: " , quality_count_form , "counts " ,sep=" ") # concatenate text with paste function and then use TeX function in ggplot layer "annotate"
  
  plot <- ggplot(a_data, aes(x=a_quality)) +
    geom_vline(xintercept = 0.9, linetype = "dashed") +
    geom_histogram(fill="lightblue", binwidth = 0.05) +
    scale_x_continuous(lim = c(-0.05, NA), breaks = c(0,0.25,0.5,0.75,0.9,1.0), expand = waiver()) + 
    xlab(TeX("imputation quality ($r^{2}$) ")) +
    ylab("count") +
    annotate(geom = 'text', label = TeX(string_text) , parse=FALSE, x = -Inf , y = Inf, hjust = 0 , vjust = 1.25) + # >= in unicode: \u2265    , quality_count_form, " counts"
    theme_bw()
  return(plot)
}


imputation_quality_vs_effect_weight_scatter_func <- function(a_data, a_score, a_effect_weight, a_quality){
  
   num_var <- variants_file_raw %>% filter(score==a_score) %>% select(all_of("include")) %>% nrow()
      number_variants <- format(num_var, big.mark= ",", scientific =FALSE)
      num_var_inc <-  variants_file_raw %>% filter(score==a_score) %>% select(all_of("include")) %>% filter(include == "1") %>% nrow()
      number_variants_included <-format(num_var_inc, big.mark= ",", scientific =FALSE)
      coverage <- round(num_var_inc/num_var*100, digits = 2)
     

    # scaling x-axis to show NA-values in plot as negative r^2
    scale <- if  (number_variants != number_variants_included) {
      scale_x_continuous(lim = c(-0.05 , NA), breaks = c(0,0.25,0.5,0.75,0.9,1.0)) 
    } 
      
      else {
      scale_x_continuous(lim = c(0 , NA), breaks = c(0,0.25,0.5,0.75,0.9,1.0)) 
      } 
    
    data_color <- a_data %>% mutate(Color= ifelse(get(a_quality) >= 0, "blue", "red"))
      
      
    plot <- ggplot(data_color, aes(y = get(a_effect_weight), x = get(a_quality), color=Color)) +
    geom_vline(xintercept = 0.9, linetype = "dashed") +
    geom_point(size = 1, alpha=0.2) +
    scale_color_identity() +
    scale +
    ylab("effect weight") + xlab(TeX("imputation quality ($r^{2}$)")) + # hier wieder weiter mit labeln
    annotate(geom = 'text', label = paste0(" ", number_variants_included," of ", number_variants, " variants \n (", coverage, " %)") ,  x = -Inf , y = Inf, hjust = 0 , vjust = 1.1) + 
    theme_bw()
  return(plot)
    
}    

```

# Analysis of phenotype traits with Polygenic Scores (PGS) 

In this report a comparison between Polygenic Scores to phenotype data is created.


```{r read phenotypes and scores}
phenotype_file_raw <- read.csv(params$input_phenotypes)
PGS_file_raw <- read.csv(params$input_PGS_data)
```

```{r read variants}
variants_file_raw <- read.table(params$input_variants_file, header=TRUE, sep = ",")
```
 
# Input check
Simple check in order to be sure if the input files are correct.

Input Files:

  - Phenotype File : ***`r params$input_phenotypes` *** 
  - PGS File : ***`r params$input_PGS_data` ***
  - Variant File : ***`r params$input_variants_file` ***

  
```{r select} 
# only "sample ID (proband) " and input_phenotypes_specification (e.g. "measured height (BL_ku_groesse)") are selected and only when a height was actually measured

phenotype_file <- phenotype_file_raw %>%
  select("proband") %>%
  drop_na()

PGS_file <- PGS_file_raw %>%
  select("sample")%>%
  drop_na()
```

```{r merge samples}
# inner join only merges those rows where in BOTH files "proband" and "sample" are present
merge_samples <- inner_join(phenotype_file, PGS_file, by = c("proband" = "sample"))
```

# Check for sample intersection

```{r count complete samples}
number_complete_sample_phenotype_file <- phenotype_file_raw %>%
  nrow()
number_complete_sample_PGS_file <- PGS_file_raw %>%
  nrow()
```
  
```{r count unique samples}
# Attention: also unique values < 0 (problematic e.g with phenotype data for height --> no negative height) are included
number_sample_phenotype_file <- get_unique_number_func(phenotype_file, "proband")
number_sample_PGS_file <- get_unique_number_func(PGS_file, "sample")
```

```{r intersection of samples}
# NA deleted
number_sample_intersection <- get_unique_number_func(merge_samples, "proband")
```

<!-- 

everything from here inside loop !

-->

```{r length check}
PGS_vector <-unlist(strsplit(params$input_PGS_numbers, ","))
phenotypes_vector <-unlist(strsplit(params$input_phenotypes_specification, ","))
if ((length(PGS_vector) != length(phenotypes_vector))){
    stop("ERROR: number of Phenotypes and corresponding Polygenic Scores is not the same. Try again with new input parameters!")
}
```



```{r definition of function: loop calculation of sample number, na values and intersection }
count_values_func <- function(a_scores_file, a_phenotypes_file, a_scores, a_phenotypes){
  # calculates NA-values in dataframe for each column separately and creates table with sum of NA-values printed  for each variable
  
  n = length(a_scores) # ==  length(a_phenotypes) ; because already checked earlier
  result_count = NULL
  
  for (i in 1:n){
    phenotype <- a_phenotypes[i]
    score <- a_scores[i]
  
   
    # number of unique samples
    # sample NA's excluded, but that makes no difference, because a sample-number exists if to be regareded as a patient
    number_sample_phenotype_file <- get_unique_number_func(a_phenotypes_file, "proband") # hard coded
    number_sample_PGS_file <- get_unique_number_func(a_scores_file, "sample") # hard coded
    
    # number of NA values
    phenotype_na_count <- sum(is.na(a_phenotypes_file %>%select(all_of(phenotype))))
    score_na_count <- sum(is.na(a_scores_file %>%select(all_of(score))))
    
    
    # number of samples after intersection
     merge_single <- inner_join(a_phenotypes_file %>% select(all_of(c("proband", phenotype))), a_scores_file %>% select(all_of(c("sample", score))), by = c("proband" = "sample"))
     number_sample_intersection <- get_unique_number_func(merge_single, "proband") # hard coded
   
     # number of sample intersection without NA values! 
     merge_intersection_without_NA <- inner_join(a_phenotypes_file %>% select(all_of(c("proband", phenotype))), a_scores_file %>% select(all_of(c("sample", score))), by = c("proband" = "sample")) %>% drop_na()
     number_merge_intersection_without_NA <- get_unique_number_func(merge_intersection_without_NA, "proband")
     
     # create table with calculated values 
    result_count <- rbind(result_count, c(phenotype = phenotype, number_sample_phenotype_file = number_sample_phenotype_file, phenotype_na_count = phenotype_na_count, score = score, number_sample_score_file = number_sample_PGS_file ,score_na_count = score_na_count, sample_intersection = number_sample_intersection, sample_intersection_without_NA = number_merge_intersection_without_NA ))
    
    
  }
  return(result_count)
}
```


```{r count values}
datatable(count_values_func(PGS_file_raw, phenotype_file_raw, PGS_vector, phenotypes_vector))
```


```{r selection of score and r2  variants file}
filter_imputation <- variants_file_raw %>% select(all_of(c("score","r2"))) %>% drop_na()

```

```{r selection of score, r2 and effect weight variants file}
filter_effect_vs_quality <- variants_file_raw %>% select(all_of(c("score","r2", "effect_weight"))) %>% replace_na(list(r2 = -0.05)) # drop_na
```




# Plots for every Phenotype - PGS pair


```{r plots,  echo=FALSE, include=TRUE, results='asis', warning=FALSE}
plot_loop <- function(a_scores_file, a_phenotypes_file, a_scores, a_phenotypes, a_variants_file){
    
    n = length(a_scores) # ==  length(a_phenotypes) ; checked already
    for (i in 1:n){
      phenotype <- a_phenotypes[i]
      score <- a_scores[i]
      
      # filter corresponding phenotype-pgs pair
      filter_phenotype <- a_phenotypes_file %>% select(all_of(c("proband", phenotype))) %>% drop_na() 
      filter_score <- a_scores_file %>% select(all_of(c("sample", score)))%>% drop_na()
      
      # delete all phenotype values < 0 --> negative height, BMI etc. makes no sense!
      if (params$input_remove_neg_values == TRUE){
        filter_phenotype <- filter_phenotype %>% filter(get(phenotype) >= 0) 
      # for my phenotypes all values are positive! e.g Lpa concentration, height etc. --> negative values are useless and wrong 
      }
      merge_single <- inner_join(filter_phenotype, filter_score, by = c("proband" = "sample"))
      
      # filter only specific PGSxxx  and save second column (=="r2") in quality --> simpel solution (hard coded)
      filter_imputation_per_PGS <- filter_imputation %>% filter(score==a_scores[i])
      
      # filter specific PGSxxx and make  effect vs quality plot!
      filter_effect_vs_quality_per_PGS <- filter_effect_vs_quality %>% filter(score==a_scores[i])
      
      # calculation of coverage per PGSxxx
      num_var <- variants_file_raw %>% filter(score==a_scores[i]) %>% select(all_of("include")) %>% nrow()
      number_variants <- format(num_var, big.mark= ",", scientific =FALSE)

      num_var_inc <-  variants_file_raw %>% filter(score==a_scores[i]) %>% select(all_of("include")) %>% filter(include == "1") %>% nrow()
      number_variants_included <-format(num_var_inc, big.mark= ",", scientific =FALSE)
     
      coverage <- round(num_var_inc/num_var*100, digits = 2)
      
      # calculation of phenotype mean
      calculated_phenotype_mean <- mean(a_phenotypes_file[[phenotype]], na.rm = TRUE)
      calculated_phenotype_mean_form <- format(calculated_phenotype_mean, big.mark= ",", scientific =FALSE)
    
      
    
      
      # create Markdown Headlines in loop 
      cat(paste0("\n\n## ", phenotype, " - ", score, "\n\n mean = ",calculated_phenotype_mean_form )) 
      # need linebreaks \n\n in order to work properly, also important: results == 'asis' in chunk description of this cell!
      
      # plotting 
      p1 <- scatterplot_func(merge_single, score, phenotype)
      p2 <- plot_pgs_by_decile(merge_single, score, phenotype, quantile = 0.1)
      p3 <- imputation_quality_histogram_func(filter_imputation_per_PGS, filter_imputation_per_PGS$r2)
      p4 <- imputation_quality_vs_effect_weight_scatter_func(filter_effect_vs_quality_per_PGS, a_scores[i], "effect_weight", "r2") # hard coded
      p1234 <- p1 + p2 + p3 + p4 
      # + plot_annotation(title = score, phenotype,  caption= paste0("caption ", phenotype)) 
      
      print(p1234)
    }
    
}

plot_loop(PGS_file_raw, phenotype_file_raw, PGS_vector, phenotypes_vector, variants_file_raw)
```





